\section{backgroud}
\label{background}


\subsection{Temporal Point Processes and Order Book Modeling}
\Ac{tpp} model event arrivals in continuous time via a conditional intensity $\lambda^\ast(t)$~\cite{Shchur2021NeuralReview}. Marked TPPs (MTPPs)~\cite{Upadhyay2018DeepProcesses} enrich this framework by attaching categorical marks to each event, capturing both timing and type. In financial markets, MTPPs are widely applied to model limit order book dynamics, where marks naturally encode order types (e.g., limit, cancel, market). A prominent example is the Hawkes process, a self-exciting MTPP extensively used for order book modeling due to its ability to capture clustering in order arrivals~\cite{Jain2024LimitProcess, Jain2025NoBooks}. The neural Hawkes process~\cite{Mei2017TheProcess} extends this idea with neural architectures that learn non-linear, history-dependent intensities, outperforming classical parametric forms. However, these models assume a single event per timestamp, an unrealistic constraint in limit order books where multiple actions often co-occur. To overcome this, Chang et al.~\cite{Chang2023} introduce the set-valued MTPP (Set-MTPP), which generalizes marks from singletons to sets, enabling principled modeling of simultaneous market events. This foundation is critical for capturing realistic order flow and underpins our anomaly detection framework. 






% \subsection{Market Participants and Manipulators}


% Normally, in centralized exchanges, order flows are primarily generated by two types of participants: \textbf{investors}, who submit directional trades, and \textbf{market makers}, who continuously provide liquidity by quoting both sides of the book while managing inventory risk~\cite{GaneshReinforcementMarket}. Under this normal market composition, the aggregate flow admits a stable probabilistic structure. However, if a \textbf{manipulator} enters the market and intends to influence the market through pump-and-dump and wash trading, etc. To impact the market, the manipulators submit their strategic orders into the natural dynamics of investors and market makers. 

\subsection{Market Participants and Manipulators}

In centralized exchanges and under normal trading conditions, order flows predominantly arise from two types of participants: investors, who submit directional orders to express market views, and market makers, who continuously quote on both sides of the book to provide liquidity while managing inventory risk~\cite{GaneshReinforcementMarket}. Under this natural market composition, the aggregate order flow exhibits a relatively stable probabilistic structure. However, when \textbf{manipulators} intervene—through practices such as pump-and-dump schemes or wash trading—they deliberately inject strategic orders into this equilibrium, distorting the underlying dynamics generated by investors and market makers and undermining the integrity of the market.

eqwe
\subsection{Anomaly Detection}
1. deviation from the model. First order moment, second order moment
2. Querying 

% \subsection{Anomaly Detection in Event Stream or Flow}
% 

