\begin{abstract}
Simulating limit-order-book (LOB) order flow is a fundamental generative task for
market-making world models, agent training, and counterfactual market analysis.
Neural marked temporal point processes (MTPPs) predict the next LOB event
accurately, but their simulation behaviour is largely unexamined: closed-loop, the
learned self-excitation drifts or explodes, and we show that the standard remedy
--- rescaling the intensity to match the real event rate --- often has \emph{no
solution} for unbounded models. To close this gap we propose \ssppp{} (Stable
State-Space Point Process), which keeps the expressive state-space backbone of
\sppp{} verbatim and replaces only the output heads: a \emph{softmin-bounded rate
head} whose intensity carries a hard closed-form ceiling (an exact dominating rate
for Ogata thinning) with a floor of exactly zero, and a \emph{rate-neutral}
soft-max mark head that can be arbitrarily deep without perturbing the total rate.
Extensive experiments on three Coinbase assets (BTC, ETH, SOL; production rates of
$24$--$48$ events/s), against five baselines with three training seeds each,
lossless streaming evaluation, and \emph{verified} rate calibration for every
model, show that \ssppp{} is the strongest predictor on two of three assets while
remaining verifiably calibratable closed-loop with competitive simulated
structure. Treating calibratability itself as an experimental outcome yields the
sharpest result: the unbounded parent fails verified calibration on 7 of 9
checkpoints --- on one asset losing every usable simulator --- with a
near-discontinuous rate response characteristic of criticality, while the bounded
head verifies on all nine. A per-gap hazard decomposition prices the residual
prediction gap exactly where the design puts it: quiet-regime hazard, the
deliberate cost of the bound. The full recipe trains with an exact parallel
prefix scan and simulates by exact-ceiling thinning with $O(1)$ carried state,
at verified real rates.
\end{abstract}

\noindent\textbf{Code} --- all tables and figures in this paper regenerate
end-to-end from collection and plotting scripts, released upon publication.
