\section{Background and Related Work}
\label{sec:background}

\paragraph{Marked temporal point processes.}
An MTPP models a sequence $\mathcal{H}_T=\{(t_i,e_i)\}_{i=1}^{N_T}$ of events with
times $t_i$ and types $e_i\in\mathcal{E}$, a finite set of atomic event types, through
per-type conditional intensities
$\lambda_e(t\mid\mathcal{H}_t)$, conditioned on the history $\mathcal{H}_t$ of all
events up to time $t$ (model parameters $\theta$ suppressed; we abbreviate
$\lambda_e(t)$ when the conditioning is clear). Writing
$\lambda(t)=\sum_{e\in\mathcal{E}}\lambda_e(t)$ for the total (ground)
intensity and $p(e\mid t)=\lambda_e(t)/\lambda(t)$ for the mark distribution,
the log-likelihood of a sequence is
\[
\log\mathcal{L}
=\sum_i \Bigl[\log\lambda(t_i)+\log p(e_i\mid t_i)\Bigr]
-\int_0^T\!\lambda(u\mid\mathcal{H}_u)\,du ,
\]
the integral being the \emph{compensator}
$\Lambda(t)=\int_0^t\lambda(u\mid\mathcal{H}_u)\,du$. By the random time change
theorem, the
rescaled masses $u_i=\Lambda(t_i)$, $\Delta u_i = u_i-u_{i-1}$ are i.i.d.\
$\mathrm{Exp}(1)$ iff the intensity is correctly specified --- we report their mean
(target 1) and KS distance as timing-calibration diagnostics. Sampling uses Ogata
thinning, which requires a dominating rate $\bar\lambda\ge\lambda(t)$; a model
whose intensity has no computable upper bound must resort to heuristic (and
bias-prone) dominating-rate estimates.

\paragraph{The \sppp{} backbone.}
\sppp{} \citep{chang2025ctssm,Shi2022StateSimulation} is a state-space point
process: $L$ stacked latent-linear-Hawkes/SSM layers with diagonal state,
ZOH-discretized between events and computable by parallel scan,
\[
x^{(\ell)}_{t'}=\bar A^{(\ell)} x^{(\ell)}_{t}
+(\bar A^{(\ell)}\!-\!I)\,\tilde B^{(\ell)} h^{(\ell-1)}
\;\bigl[+\,\tilde E^{(\ell)}\alpha\ \text{at an event}\bigr],
\]
with a LayerNorm'd residual readout
$h^{(\ell)}=\mathrm{LayerNorm}\bigl(\mathrm{GELU}(y^{(\ell)})+h^{(\ell-1)}\bigr)$
producing the continuous-time hidden representation $h_t:=h^{(L)}(t)\in\mathbb{R}^{H}$. The published model
reads the rate from $h_t$ through a projection-and-softplus head,
$\lambda = \mathrm{softplus}(w^{\!\top}h_t+b)$, which is unbounded above and has no
learnable scale; marks are coupled to the same rate field (our benchmark
implementation of this baseline splits $h_t$ into disjoint halves for the rate
and mark heads). \ssppp{} keeps everything up to and including $h_t$ and changes
only the two heads.

\paragraph{Evaluation axes.}
We score \emph{prediction} per genuine test event (NLL splits, type accuracy,
expected-time MAE, time-rescaling diagnostics) and \emph{simulation} by
calibrated closed-loop roll-outs scored on stylized facts of real order flow
\citep{cont2001empirical,vyetrenko2020getreal}; the protocol is detailed in
Section~\ref{sec:experiments}.

\paragraph{Neural TPPs for event streams.}
Neural MTPPs learn history-dependent intensities with recurrent
\citep{du2016recurrent,Mei2017TheProcess}, attention-based
\citep{zhang2020self,zuo2020thp}, jump-diffusion \citep{zhang2024njdtpp}, and
continuous-time state-space \citep{chang2025ctssm} encoders
\citep[survey:][]{Shchur2021NeuralReview}. \citet{Shi2022StateSimulation}
specialize neural Hawkes models to LOB event streams and evaluate both prediction
and simulation. Notably, the state-space line is evaluated almost exclusively
teacher-forced: \citet{chang2025ctssm} report state-of-the-art held-out likelihood
but never roll the model out (they avoid thinning altogether) --- closed-loop
simulation is an unaddressed gap of the \sppp{} paper, and our experiments show it
is not benign: the unbounded head is frequently \emph{uncalibratable} in free
roll-out (Table~\ref{tab:caloutcomes}). Our contribution is
orthogonal: we keep the backbone verbatim and change only what is read
\emph{out} of it.

\paragraph{Hawkes modeling of order flow.}
Parametric Hawkes processes are the classical LOB event model
\citep{Hawkes1971,Hawkes2018HawkesReview,Jain2024LimitProcess,jain2024review}, with
state-dependent extensions \citep{Morariu-Patrichi2022State-dependentModelling}.
Their linear structure gives exact stationarity certificates but limited
expressiveness; \ssppp{} takes the complementary route --- full neural
expressiveness with stability restored by a closed-form bound. In the neural
TPP literature such bounds appear mainly as computational devices for
thinning; here the bound is a modeling decision with an asymmetric requirement
(hard ceiling, exactly-zero floor), validated by our calibration study.

\paragraph{Generative LOB simulators.}
Learned order-flow generators span GANs \citep{li2020stockgan,coletta2021lobgan,
coletta2022worldagent}, token-level autoregressive state-space models
\citep{nagy2023lobs5}, diffusion-guided agents \citep{huang2026diga}, and
foundation-model-scale transformers \citep{mars2025,sood2026tradefm}; agent-based
simulators \citep{byrd2020abides,frey2023jaxlob} remain the RL workhorse. These
systems are scored on distributional realism: stylized facts
\citep{cont2001empirical,vyetrenko2020getreal} and the KS/Wasserstein suites of
LOB-Bench \citep{nagy2025lobbench}, whose authors report error accumulation over
long roll-outs. \ssppp{} differs on two counts: one likelihood-based model is
held to \emph{both} prediction and closed-loop realism, and the simulated rate
is calibrated and \emph{verified} --- a closed-loop stability certificate no
generative simulator above carries.

