\section{Discussion and Conclusion}
\label{sec:conclusion}

We held six neural MTPPs to both sides of the prediction--simulation tension
on three Coinbase assets at production event rates ($24$--$48$ events/s),
under a protocol in which failures are visible and verified
rate-calibration failures count as results. The
state-of-the-art predictor \sppp{} is not a usable simulator: its roll-out
rate runs away on 7 of 9 checkpoints, and on SOL no usable checkpoint
remains. \ssppp{}, which differs from it only in the output heads, wins
every prediction metric on SOL, leads on ETH, ties the strongest baseline on
BTC marks, verifies rate calibration on all nine checkpoints, and leads the stable
baselines on most of the twelve stylized facts.

Two conclusions follow. First, closed-loop stability is a design property
of the intensity head, not a training outcome. The unbounded head's failure
carries a reproducible critical-point signature, a closed-loop rate
near-discontinuous in the rescaling constant, so no post-hoc rescaling can
recover it; the softmin bound removes the failure at the source and turns
calibration into a verified one-dimensional root-find. Second, the bound is
nearly free: it costs no predictive accuracy on any asset, trains at scan
speed like its parent, and its closed-form ceiling buys exact Ogata thinning
at throughput far above real time.

The evidence has limits: three streams from a single crypto venue,
seven days each, and the simulator emits event types and times
rather than full book states, so realism is scored on order-flow
statistics, not price paths. The certificate guarantees a bounded rate, not realism: the
LSTM is stable yet simulates a near-Poisson stream, and even \ssppp{}
leaves a dispersion gap on the Fano factor; the remaining prediction
headroom is BTC timing, where \nhp{} leads.

Because the mark head is rate-neutral, conditioning it on exogenous state
such as an agent's resting quotes cannot destabilize the rate, so \ssppp{}
can serve as an environment model for market-making world models. The
same factorized intensity supports the importance-sampling queries of
\citet{chang2024probabilistic}, such as fill probability and
adverse-selection risk; both extensions are future work. All tables and
figures are regenerated by the released scripts.
