\section{Experiments}
\label{sec:experiments}

\subsection{Setup and validity}
\label{sec:setup}
All models are trained and evaluated identically: Gemini ETH-USD event-driven order
flow (62 event types, single-item categorical marks, real rate $\approx 2.32$
events/s), sequence length 64 / stride 32, 40 epochs, hidden size 64, one A40 GPU per
model. The zoo covers seven models: \nhp{} (the Mei--Eisner neural Hawkes CT-LSTM;
\emph{hawkes} and \emph{ct-lstm} are the same class and reported once), an LSTM
decoder and a SAHP-style causal-attention decoder \citep{zhang2020self} --- both
\emph{adapted generic backbones} over our shared input embedding, not paper-faithful
re-implementations --- PCT-LSTM (a per-type parallel CT-LSTM), \sppp{}, and
\ssppp{} variants. Three implementation bugs found in the vendored baselines --- a
one-event-stale state anchor in the query path, the neural-Hawkes decay applied with
the previous inter-event gap, and detached channel-embedding gradients --- were fixed
and all affected baselines retrained; models whose code was untouched reproduced
their previous results bit-identically (\sppp{}, PCT-LSTM, \ssppp{}-G1), validating
the rerun.

\subsection{Prediction}
\label{sec:prediction}
Table~\ref{tab:prediction} reports per-genuine-event metrics: \emph{overall} = time +
mark NLL, the time-rescaling calibration (mean rescaled mass, target 1; KS distance
from Exp(1)), time-MAE, type accuracy, and perplexity.

\begin{table}[t]
\centering\small
\setlength{\tabcolsep}{3.2pt}
\begin{tabular}{lcccccc}
\toprule
model & overall$\downarrow$ & KS$\downarrow$ & mean\,$u$ & tMAE & ACC$\uparrow$ & PPL$\downarrow$\\
\midrule
\nhp{} (CT-LSTM) & \textbf{0.667} & 0.235 & 1.91 & \textbf{0.300} & \textbf{0.379} & \textbf{9.01}\\
\textbf{\ssppp{}-softmin} & 0.982 & \textbf{0.189} & \textbf{1.84} & 0.384 & 0.331 & 10.12\\
\ssppp{}-G1 & 1.078 & 0.253 & 2.09 & 0.344 & 0.343 & 9.95\\
\ssppp{}-lsinit & 1.149 & 0.237 & 2.18 & 0.329 & 0.337 & 10.09\\
PCT-LSTM & 1.327 & \textbf{0.189} & 1.87 & 0.397 & 0.276 & 12.70\\
SAHP & 1.524 & 0.448 & 0.99 & 0.435 & 0.356 & 9.97\\
LSTM & 1.593 & 0.485 & 0.78 & 0.464 & 0.343 & 10.43\\
\sppp{} & 2.959 & 0.437 & 4.29 & 1.430 & 0.326 & 10.45\\
\bottomrule
\end{tabular}
\caption{Prediction per genuine test event. \nhp{} is the best predictor across the
board; \ssppp{}-softmin is the best of everything else and holds the best timing
\emph{calibration} in the table, and every \ssppp{} variant beats its unbounded
\sppp{} parent by $\ge$1.8 nats.}
\label{tab:prediction}
\end{table}

\subsection{Simulation}
\label{sec:simulation}
Table~\ref{tab:simulation} scores 600\,s closed-loop roll-outs by relative error
against real data on the stylized-facts set.

\begin{table}[t]
\centering\footnotesize
\setlength{\tabcolsep}{2.6pt}
\begin{tabular}{lccccc}
\toprule
model & rate (real 2.32) & rate\_re$\downarrow$ & Fano\_re$\downarrow$ & clus\_re$\downarrow$ & rACF\_re$\downarrow$\\
\midrule
SAHP & 1.76 & \textbf{0.24} & 0.29 & \textbf{0.08} & 0.38\\
LSTM & 1.20 & 0.49 & 0.46 & 0.28 & \textbf{0.23}\\
\ssppp{}-G1 & 52.7 & 21.7 & 1.02 & 0.22 & 0.40\\
\ssppp{}-lsinit & 78.4 & 32.8 & \textbf{0.27} & 1.06 & 0.54\\
\textbf{\ssppp{}-softmin} & \textbf{32.3} & \textbf{12.9} & 1.97 & 3.69 & 3.52\\
PCT-LSTM & 23.6 & 9.2 & 0.34 & 0.32 & 1.11\\
\nhp{} (CT-LSTM) & 66.4 & 27.6 & 6.63 & 3.33 & 0.24\\
\sppp{} & 87.3 & 36.6 & 11.63 & 2.60 & 0.00\\
\bottomrule
\end{tabular}
\caption{Closed-loop stylized-facts fit (600\,s roll-outs; rel-err vs real). The
mirror image of prediction: \nhp{}'s self-excitation runs away (66\,ev/s, Fano
6.6$\times$ off); the bounded \ssppp{} family stays contained, and softmin's open
floor gives the best rate among intensity models. The trivial LSTM/SAHP samplers
match unconditional facts best but are the weakest predictors.}
\label{tab:simulation}
\end{table}

Together the two tables trace the expressivity--stability frontier: \nhp{} takes the
expressive end (best prediction, unusable simulation), \sppp{} falls off both ends,
and \ssppp{}-softmin is the current best point --- within 0.3 nats of \nhp{} with a
provably bounded roll-out. One cross-cutting bias remains: \emph{all} intensity
models over-produce events free-running, consistent with their mean rescaled masses
$u>1$ in Table~\ref{tab:prediction} --- the endpoint-rule compensator plus windowed
training systematically underestimate the integral of the intensity (see the
ablation chain below).

\subsection{The ablation chain: locating the quiet-regime gap}
\label{sec:ablation}
We bucket each test event by trailing activity (rate over the last 8 gaps) and
measure the per-event time-NLL deficit against \nhp{}
(Table~\ref{tab:ablation}).

\begin{table}[t]
\centering\small
\begin{tabular}{lccccc}
\toprule
deficit vs \nhp{} (nats) & Q1 quiet & Q2 & Q3 & Q4 & Q5 bursts\\
\midrule
\ssppp{}-G1 (sandwich) & $+1.13$ & $+0.56$ & $+0.37$ & $+0.09$ & $-0.01$\\
\ssppp{}-lsinit & $+1.16$ & $+0.62$ & $+0.40$ & $+0.18$ & $+0.04$\\
\textbf{\ssppp{}-softmin} & $\bm{+0.68}$ & $\bm{+0.33}$ & $\bm{+0.19}$ & $\bm{+0.11}$ & $\bm{-0.02}$\\
\bottomrule
\end{tabular}
\caption{Per-event time-NLL deficit vs \nhp{}, bucketed by trailing activity
(2{,}957 events). In bursts \ssppp{} already matches \nhp{}; the deficit is a
quiet-regime phenomenon, cut 41\% by the open-floor softmin head.}
\label{tab:ablation}
\end{table}

\paragraph{Diagnosis.} The G1 head's floor $\ell_-\approx0.36$\,ev/s bleeds
compensator mass $(\ell_--\lambda^*)\Delta t\approx0.33\,\Delta t$ nats in quiet
gaps --- the entire Q1 deficit; in bursts \ssppp{} already matched \nhp{}.

\paragraph{Step 1 --- lsinit.} An S4-style log-spaced decay initialization
(timescales $0.02$--$60$\,s) retains $\sim$30\,s modes after training and optimizes
better (val $-18.1$ vs $-17.0$), but the deficit is flat --- the decay spectrum was
not the binding constraint.

\paragraph{Step 2 --- softmin head.} Replacing the sandwich with the one-sided cap
(Eq.~\ref{eq:softmin}) keeps the exact thinning ceiling and opens the floor. The
realized intensity floor drops $0.34\!\to\!0.11$, the $\lambda(\delta)$ relaxation
plateau at $0.36$ is broken ($0.19$ at 5\,s, $0.12$ at 20\,s), the Q1 deficit falls
41\%, and the overall NLL improves $1.08\to0.98$ --- while the \emph{training} loss
gets worse ($-12.9$ vs $-18.1$): the old floor was partly exploiting the biased
endpoint compensator rather than fitting the data.

\paragraph{What remains.} The residual gap (0.20 nats timing + 0.12 marks) traces to
the frozen ZOH input hold: between events the state converges to
$x_\infty=\tilde B u_{\text{held}}$ ($\|x\|\!\approx\!30$ at 60\,s) rather than to
zero, keeping the quiet-regime rate elevated. A zero-asymptote ablation reaches
\nhp{}-level quiet ($\lambda\approx0.02$); a \emph{leaky hold}
$u_{\text{held}}(\delta)=e^{-\rho\delta}u_{\text{held}}$ is left as future work.

\subsection{From bounded predictor to calibrated simulator}
\label{sec:arc}
We now resolve the remaining prediction--simulation gap through single-factor
ablations in a long-context regime (sequence length 1024, non-overlapping windows
where stateful training requires it; test-slice real rate 3.48\,ev/s; all roll-outs
600\,s $\times$ 32 sequences, carried-state for the \sppp{} family). Numbers in this
subsection are not comparable to Tables~\ref{tab:prediction}--\ref{tab:ablation}
(different context length, stride, and training-step budget).

\paragraph{Roll-out memory unlocks long-range structure.} With carried-state
simulation, the volatility long-memory fact moves for the first time in this study:
\sppp's simulated $|r|$-autocorrelation (F6) reaches $0.040$ against a real $0.056$
--- every sliding-window protocol before it was capped near noise ($\le 0.03$),
because a 64-event window spans only $\sim$27\,s of market time. Memory truncation,
not model class, was the binding constraint.

\paragraph{Stateful training owns structure and prediction --- not the rate.}
Pairing cold-start against TBPTT training, everything else identical:

\begin{table}[t]
\centering\small
\setlength{\tabcolsep}{2.7pt}
\begin{tabular}{lcccccc}
\toprule
 & overall$\downarrow$ & tMAE(s) & mean\,$u$ & rate & clus\_re$\downarrow$ & rACF\_re$\downarrow$\\
\midrule
\sppp{} cold & 1.386 & 15.61 & 1.91 & 27.6 & 14.9 & 26.8\\
\sppp{} TBPTT & \textbf{1.295} & \textbf{2.65} & 1.86 & 28.0 & \textbf{0.12} & \textbf{0.43}\\
\ssppp{} cold & 1.672 & 0.49 & 2.20 & 22.4 & 9.8 & 17.6\\
\ssppp{} TBPTT & \textbf{1.428} & 0.49 & 1.96 & 23.6 & 8.2 & 9.3\\
\bottomrule
\end{tabular}
\caption{TBPTT ablation (endpoint compensator in all arms; rate in ev/s vs real
3.48; clus/rACF = rel-err of volatility clustering and $|r|$-ACF vs real). Stateful
training improves prediction across every metric and transforms simulated temporal
structure, while leaving the free-running rate and per-gap mass (mean $u$)
essentially unchanged --- the rate is not a state-regime problem.}
\label{tab:tbptt}
\end{table}

Prediction improves across the board (\sppp's pathological timing collapses from
15.6\,s to 2.65\,s MAE) and simulated clustering reaches near-real levels
(Table~\ref{tab:tbptt}) --- but the roll-out still runs $\sim$7$\times$ hot with
mean $u\approx1.9$: the remaining inflation is a property of the \emph{loss}, not of
the states.

\paragraph{The unbiased compensator fails --- for optimization, not statistical,
reasons.} Replacing the endpoint rule with the Mei--Eisner global Monte-Carlo
compensator (32 uniform samples per window) collapses both models to near-silence
(free-run 0.05--0.08\,ev/s; mean $u$ overshoots from 2.2 to 0.35; teacher-forced NLL
degrades by 5+ nats) while its training loss decreases smoothly throughout. A
diagnostic on the trained control confirms the estimator's expectation is correct
(MC/endpoint $=6.4\times$, matching the model's true intensity mass); the failure is
dynamical: the estimator's heavy-tailed penalty (2$\times$ draw-to-draw swings) is
systematically trimmed by gradient clipping under Adam, and maximum likelihood learns
to exploit the sparse audit --- spiky intensity at events (rarely sampled, since
samples weight by time), crushed baseline elsewhere (audited every step). Unbiased in
expectation is not unbiased under clipped adaptive optimization; a stratified
per-gap estimator (one sample per gap: every gap audited, variance from within-gap
variation only) is the indicated repair.

\paragraph{Post-hoc calibration closes the rate --- and improves structure.}
Applying the certificate-preserving calibration of the method section to the trained
\ssppp{} checkpoints (both converge in 6 probes to $k=0.386$: the apparent
7$\times$ rate inflation corresponds to a 2.6$\times$ over-scaled head, the rest
being closed-loop amplification):

\begin{table}[t]
\centering\small
\setlength{\tabcolsep}{3.4pt}
\begin{tabular}{lccccc}
\toprule
\ssppp{} arm & rate (real 3.48) & rate\_re & Fano\_re & clus\_re & rACF\_re\\
\midrule
cold, uncal. & 22.4 & 5.43 & 0.73 & 9.76 & 17.61\\
cold, cal. & 3.17 & \textbf{0.09} & 0.35 & 2.87 & 5.72\\
TBPTT, uncal. & 23.6 & 5.79 & 0.64 & 8.20 & 9.32\\
TBPTT, cal. & \textbf{2.59} & 0.26 & \textbf{0.35} & \textbf{0.45} & \textbf{0.28}\\
\bottomrule
\end{tabular}
\caption{Rate-calibration ablation (simulation-only, no retraining). Calibration
improves \emph{every} structure fact --- the apparent structure deficit of
over-firing simulators is largely a clock-rate artifact (facts computed on a
7$\times$-hot stream have mis-scaled buckets) --- and the TBPTT training advantage
persists after calibration.}
\label{tab:calibration}
\end{table}

Two effects separate cleanly (Table~\ref{tab:calibration}): calibration fixes the
rate by construction and repairs the bucket-scale distortion of every fact, and
stateful training contributes structure that calibration alone does not (clustering
0.45 vs.\ 2.87 within the calibrated pair). The combined recipe --- \textbf{\ssppp{}
heads + TBPTT training + carried-state roll-out + post-hoc calibration} --- is the
configuration we propose: near-real event rate, close-to-empirical clustering and
return-autocorrelation, a hard closed-form rate ceiling, and exact thinning
throughout. The calibrated model is no longer the maximum-likelihood point (the knob
trades likelihood for simulation fidelity), and the residual $\sim$25\% rate gap
reflects probe-versus-full-horizon drift; the principled endpoint --- emergent
calibration via a low-variance unbiased compensator on stateful training --- remains
open and is the natural next experiment.
