\section{Introduction}
\label{sec:intro}

The limit order book (LOB) is the core mechanism of electronic markets: a
continuous-time stream of order placements, cancellations, and trades whose fine
structure drives price discovery and liquidity. Two tasks are asked of a generative
model of this stream. The first is \emph{prediction}: given the event history, score
the time and type of the next event. The second is \emph{simulation}: roll the model
forward freely, feeding its own samples back as history, and reproduce the stylized
facts of real order flow --- calibrated event rates, burst--lull clustering, and
heavy-tailed returns \citep{cont2001empirical}. Simulation is the harder and, for
downstream use (market-making world models, agent training, counterfactual
analysis), the more valuable of the two \citep{mars2025,nagy2025lobbench}.

Neural MTPPs \citep{du2016recurrent,Mei2017TheProcess,Shchur2021NeuralReview} excel
at prediction, and continuous-time state-space variants
\citep{chang2025ctssm,Shi2022StateSimulation} push held-out likelihood to the state
of the art. But that literature evaluates almost exclusively teacher-forced, and the
very self-excitation that wins the one-step likelihood becomes a liability
closed-loop: each sampled event raises the intensity, which produces more events,
and with no mechanism bounding the rate the roll-out drifts hot or explodes. The
standard patch --- rescale the intensity until the simulated rate matches reality
--- turns out to be unavailable exactly where it is needed most: we find trained
unbounded state-space checkpoints whose closed-loop rate is \emph{near-discontinuous}
in the rescaling constant (a $10^{-3}$ change flips the rate across the entire
tolerance band) so that no constant rescaling exists, and the failure worsens
systematically with the event rate of the asset. The field thus faces an
\emph{expressivity--stability frontier}: expressive intensities predict but cannot
be trusted (or even calibrated) closed-loop, while trivially stable samplers
simulate innocuously but predict poorly.

\ssppp{} is a head transplant, not a new backbone. We keep the stacked
latent-linear-Hawkes state-space backbone verbatim --- its diagonal
zero-order-hold dynamics, parallel-scan computability, and LayerNorm'd stack
output $u(t)$ --- and replace only what comes after $u(t)$:
\begin{itemize}
\item a \textbf{softmin-bounded rate head}: a gated bounded state
  $h=\sigma(W_o u)\odot\tanh(u)$ is read out through a smooth one-sided cap
  $z = c-\mathrm{softplus}(c-w^{\!\top}h-b)$, so
  $\lambda = s\cdot\mathrm{softplus}(z) \le s\cdot\mathrm{softplus}(c)$ is a
  \emph{hard closed-form ceiling} --- an exact dominating rate for Ogata thinning ---
  while $z\to z_{\text{raw}}$ as $z_{\text{raw}}\to-\infty$, so the floor is exactly
  zero and log-intensity keeps unit gradient in the quiet regime;
\item a \textbf{rate-neutral mark head}: $p^*(k\mid t)=\mathrm{softmax}(\mathrm{MLP}(u))_k$
  lives on the probability simplex, so $\sum_k \lambda_k = \lambda$ holds regardless
  of how nonlinear the mark network is --- expressiveness goes where it cannot
  destabilize the rate.
\end{itemize}

Because the backbone is identical, differences against the unbounded parent
(\sppp) are attributable to the heads. Our contributions:
\begin{enumerate}
\item \textbf{The \ssppp{} head pair} (softmin-bounded rate $\times$ rate-neutral
  marks): a provably bounded, thinning-exact intensity on an unmodified
  state-space backbone, trained with stateful (TBPTT) likelihood and simulated
  with $O(1)$-per-event carried state.
\item \textbf{A strict two-axis benchmark} on four crypto assets (Gemini ETH;
  Coinbase BTC/ETH/SOL) $\times$ six models $\times$ three training seeds:
  prediction scored losslessly on \emph{every} test event by a streaming
  evaluator with carried state, simulation scored on calibrated carried-state
  roll-outs against equal-duration real segments, checkpoint-level confidence
  intervals throughout. \ssppp{} is the best-calibrated simulator on the home
  asset and \emph{overtakes} the best predictor (\nhp) as streams grow faster,
  winning every prediction metric on SOL. A per-gap decomposition explains the
  remaining prediction gap mechanistically: it is quiet-regime hazard, the
  deliberate price of the bound.
\item \textbf{Calibratability as a measurable model property.} Every model is
  rate-calibrated by a mark-preserving bisection with \emph{full-scale
  verification}, and the outcome is reported as a result: the unbounded head
  fails verified calibration on 8/12 checkpoints with a critical-point
  signature, degrading with event rate (all three seeds on SOL); the bounded
  head fails 1/12. To our knowledge this is the first cross-asset
  quantification of closed-loop calibratability for neural TPPs.
\item \textbf{Exact post-hoc rate calibration} for factorized heads: the rate is
  linear in a learnable scale, the thinning ceiling scales with it, and the mark
  distribution is untouched --- a certificate-preserving one-dimensional
  root-find that also \emph{improves} every structure fact by repairing
  clock-rate distortion.
\item \textbf{A cautionary negative result on unbiased compensators.} The
  global Monte-Carlo compensator, though verifiably unbiased, collapses training
  under gradient clipping and adaptive optimization. Estimator variance, not
  bias alone, decides whether likelihood-level rate calibration succeeds.
\end{enumerate}
All tables and figures in this paper are generated end-to-end by released
collection and plotting scripts.
