\section{Method}
\label{sec:method}

\subsection{Stationary-Mean Calibration for Set-Valued Hawkes Ground Intensity}
\label{subsec:hawkes_mean_calibration}

We model order-book event streams as a sequence of set-valued continuous-time events
\[
\mathcal{H}_T=\{(t_i,x_i)\}_{i=1}^{N_T},
\qquad
x_i \in \mathcal{X}:=\mathcal{P}(\{1,\ldots,K\}),
\]
where $x_i$ denotes the subset of event types that occur simultaneously at time $t_i$.
Following the set-valued temporal point process formulation of Chang, Boyd, and
Smyth~\cite{chang2024probabilistic}, we separate the temporal arrival process from the
conditional set distribution. Specifically, the set-specific intensity is written as
\[
\lambda_x^*(t)=\lambda^*(t)\,p^*(x\mid t),
\qquad x\in\mathcal{X},
\]
where $\lambda^*(t)$ is the ground intensity of set-event arrivals and $p^*(x\mid t)$ is a
normalized distribution over subsets. Thus,
\[
\sum_{x\in\mathcal{X}}\lambda_x^*(t)
=\lambda^*(t)\sum_{x\in\mathcal{X}}p^*(x\mid t)
=\lambda^*(t).
\]
The set head is therefore rate-neutral: it redistributes probability mass across possible
simultaneous event sets but does not change the total arrival rate.

In the Dynamic Bernoulli version of Chang et al.'s set model, the conditional distribution
over subsets is factorized as
\[
p_\omega(x\mid t,h_\theta(t))
=\prod_{k=1}^{K}
\rho_k(t)^{\mathbf{1}(k\in x)}
\bigl(1-\rho_k(t)\bigr)^{\mathbf{1}(k\notin x)},
\]
where
\[
\rho_k(t)=\sigma\!\left(v_k^\top n(h_\theta(t))+b_k\right).
\]
Here, $h_\theta(t)$ summarizes the history before time $t$, $n(\cdot)$ is a feed-forward
network, and $\rho_k(t)$ is the conditional probability that item $k$ appears in the next
set. When set embeddings are required as recurrent inputs, Chang et al.\ represent a
non-empty set by the average of its item embeddings:
\[
e(x)=\frac{1}{|x|}\sum_{k\in x}w_k,
\qquad x\neq\emptyset,
\]
and set $e(\emptyset)=0$. This construction allows us to model simultaneous order-book
events without expanding the mark space into $2^K$ independent labels.

We now calibrate the ground intensity $\lambda^*(t)$, denoted below by $\Lambda(t)$, using a
linear Hawkes process with an exponential-mixture kernel:
\[
\Lambda(t)=\mu_0+\int_0^t \varphi(t-u)\,dN(u),
\qquad
\varphi(\tau)=\sum_{m=1}^{M}a_m e^{-\delta_m \tau}.
\]
Here, $N(t)$ counts set-event arrivals, i.e., $dN(t)$ increments once when a subset $x_i$
occurs, regardless of its cardinality. Equivalently, defining the decayed trace
\[
s_m(t)=\sum_{t_i<t} e^{-\delta_m(t-t_i)},
\]
the ground intensity can be written as $\Lambda(t)=\mu_0+\sum_{m=1}^{M}a_m s_m(t)$.

Let $\bar{\Lambda}=\mathbb{E}[\Lambda(t)]$ denote the stationary mean ground rate. Under
stationarity, $\mathbb{E}[dN(u)]=\bar{\Lambda}\,du$. Taking expectations in the Hawkes
intensity equation gives the fixed-point relation
\[
\bar{\Lambda}
=\mu_0+\int_0^\infty \varphi(\tau)\,\bar{\Lambda}\,d\tau
=\mu_0+\bar{\Lambda}\,n,
\]
where
\[
n:=\int_0^\infty \varphi(\tau)\,d\tau
=\sum_{m=1}^{M}a_m\int_0^\infty e^{-\delta_m\tau}\,d\tau
=\sum_{m=1}^{M}\frac{a_m}{\delta_m}
\]
is the branching ratio. Solving the fixed-point equation yields
\[
\bar{\Lambda}=\frac{\mu_0}{1-n},\qquad n<1.
\]
Therefore, if we set $\mu_0=R_{\mathrm{target}}(1-n)$, then
\[
\bar{\Lambda}=\frac{R_{\mathrm{target}}(1-n)}{1-n}=R_{\mathrm{target}}.
\]
The stationary mean ground rate is therefore pinned exactly to the target rate
$R_{\mathrm{target}}$, provided the process remains subcritical.

In implementation, the excitation parameters are constrained to be non-negative, e.g.,
$a_m=\operatorname{softplus}(\tilde{a}_m)$, and the branching ratio is computed on each
forward pass as
\[
n=\sum_{m=1}^{M}\frac{\operatorname{softplus}(\tilde{a}_m)}{\delta_m}.
\]
We then set $\mu_0=R_{\mathrm{target}}(1-n)$ after projecting or clipping $n$ below one.
This makes the base rate a deterministic function of the excitation mass rather than an
independently learned parameter. Consequently, the optimizer can learn the temporal
clustering shape through $a_m/\delta_m$, but it cannot inflate the long-run mean event
rate. Since the Chang-style conditional set distribution is normalized over subsets, the
set head changes only the composition of simultaneous events and leaves the scalar Hawkes
mean calibration unchanged.

\subsection{Set-Valued and Single-Item Output as a Mark-Head Choice}
\label{subsec:mark_head_choice}

A useful consequence of the factorization $\lambda_x^*(t)=\Lambda(t)\,p^*(x\mid t)$ is that
the timing model and the calibration are \emph{agnostic} to the support of the mark
distribution $p^*(\cdot\mid t)$. The ground rate $\Lambda(t)$ counts set-event arrivals
(one increment per timestamp, irrespective of cardinality), so the branching ratio $n$ and
the rate pin $\mu_0=R_{\mathrm{target}}(1-n)$ are defined entirely on the arrival clock and
are untouched by how the mark mass is distributed. The single-item versus set-valued
distinction is therefore a localized \emph{mark-head} choice over the support of $p^*$.

Write $\mathcal{X}=\mathcal{P}(\{1,\ldots,K\})$. Both heads consume the same per-type
logits $z_k=z_k(h_\theta(t))$ and differ only in the final normalization and likelihood:

\paragraph{Single-item (categorical) head.} The support is restricted to the singletons
$\{k\}$, and $p^*$ is a soft-max over types,
\[
p^*(\{k\}\mid t)=\frac{e^{z_k}}{\sum_{j=1}^{K}e^{z_j}},
\qquad
\sum_{k=1}^{K}p^*(\{k\}\mid t)=1,
\]
trained by categorical cross-entropy. Each arrival is one event type.

\paragraph{Set-valued (Dynamic Bernoulli) head.} The support is the full power set, and
$p^*$ factorizes into independent per-type Bernoullis with $\rho_k=\sigma(z_k)$,
\[
p^*(x\mid t)=\prod_{k\in x}\rho_k\prod_{k\notin x}\bigl(1-\rho_k\bigr),
\]
trained by the summed per-type Bernoulli log-likelihood. Each arrival is a subset of
event types, modelling genuine simultaneity.

\paragraph{Both heads are rate-neutral, so calibration is preserved either way.}
Rate-neutrality is the statement $\sum_{x}p^*(x\mid t)=1$ over the chosen support, which
holds in both cases: the soft-max sums to one over the $K$ singletons, while the
product-of-Bernoullis sums to one over the entire power set,
\[
\sum_{x\subseteq\{1,\ldots,K\}}\;\prod_{k\in x}\rho_k\prod_{k\notin x}(1-\rho_k)
=\prod_{k=1}^{K}\bigl(\rho_k+(1-\rho_k)\bigr)=1.
\]
In either case $\sum_{x}\lambda_x^*(t)=\Lambda(t)$, so the stationary-mean pin
$\bar{\Lambda}=R_{\mathrm{target}}$ and the branching certificate are independent of the
mark-head choice. A single event is simply the singleton $x=\{k\}$ ($|x|=1$), so the
set-valued formulation subsumes single-item output as a special case, and the multi-hot
event encoding represents it natively.

The practical guidance follows directly: pick the head to match the data, not the timing
model. When arrivals are effectively atomic, the categorical head is exact and economical;
when simultaneity ($|x|>1$) is material, the Dynamic Bernoulli head represents it
faithfully, while the categorical head must collapse each set to a single representative
type and is then a lossy approximation. Crucially, switching between the two changes only
the mark head's activation and likelihood---never the ground rate, its calibration, or the
set-valued input encoding.
