\documentclass[11pt]{article}

% --- packages -------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[numbers,sort&compress]{natbib}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\graphicspath{{figures/}}

% --- metadata -------------------------------------------------------------
\title{LGM: A Calibrated, Certifiable Factorized Neural Hawkes Process\\
       for Limit-Order-Book Simulation}
\author{Honglin Fu}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Neural marked temporal point processes (MTPPs) predict the next limit-order-book
(LOB) event well but \emph{simulate} order flow poorly: in free roll-out they
over-disperse or explode, and their stability cannot be honestly certified. We
introduce \textbf{LGM} (Linear Ground $\times$ soft-max Marks), which factorizes the
per-type intensity into a \emph{linear} scalar ground rate and a \emph{deep} soft-max
mark head. Because the marks live on the probability simplex they are rate-neutral, so
the total rate remains a pure linear Hawkes process regardless of mark-net nonlinearity.
This preserves the exact stationary-mean formula, which we \emph{pin} analytically to
solve the train/simulate calibration mismatch, and yields a gauge-free branching ratio
as a closed-form stability certificate. On Gemini ETH--USD data (62 event types), LGM is
simultaneously calibrated (free-roll rate within ${\sim}7\%$ of real), a competitive
predictor, clustered (correct Fano profile), and certifiable --- the first model in our
study to achieve all of these together.
\end{abstract}

\section{Introduction}
\label{sec:intro}
% Why simulation (not just prediction) is the goal; the two failure modes
% (calibration mismatch + gauge pathology). Cite the world-model motivation.
Order-book simulation underpins market-making research, backtesting, and reinforcement
learning of execution policies. \citep{avellaneda2008high}

\section{The LGM factorization}
\label{sec:model}
The per-type intensity factorizes into a timing rate and a mark distribution,
\begin{equation}
  \lambda_k(t) = \Lambda(t)\, p(k \mid t),
  \qquad \sum_k \lambda_k(t) = \Lambda(t),
  \label{eq:factorization}
\end{equation}
where the ground rate is a multi-timescale \emph{linear} Hawkes process and the marks
are a deep soft-max:
\begin{equation}
  \Lambda(t) = \mu_0 + \sum_{m} a_m\, s_m(t),
  \qquad
  p(\cdot \mid t) = \mathrm{softmax}\big(z_\theta(t)\big).
  \label{eq:ground}
\end{equation}
Since the soft-max lies on the simplex, $\sum_k p_k = 1$, the marks are
\emph{rate-neutral}: the total rate $\Lambda$ is unaffected by mark-net nonlinearity.
The stationary mean therefore survives in closed form,
\begin{equation}
  \bar{\Lambda} = \frac{\mu_0}{1 - n}, \qquad n = \sum_m \frac{a_m}{\beta_m},
  \label{eq:mean}
\end{equation}
and we \textbf{pin} it by setting $\mu_0 = R_{\text{target}}(1-n)$, so $\bar{\Lambda} =
R_{\text{target}}$ by construction. The branching ratio $n$ is gauge-free and serves as a
stability certificate, with $\mathrm{Fano}(\infty) = 1/(1-n)^2$.

\section{Results}
\label{sec:results}
Table~\ref{tab:headline} reports the headline comparison on Gemini ETH--USD
(62 event types, real rate ${\approx}2.38$ ev/s).

\begin{table}[t]
  \centering
  \caption{Headline comparison. LGM is uniquely calibrated \emph{and} a competitive
           predictor \emph{and} clustered \emph{and} certifiable.}
  \label{tab:headline}
  \begin{tabular}{lccccl}
    \toprule
    Model & GenAcc & Fano(1s) & branching $\rho$ & free-roll rate & sim status \\
    \midrule
    Compound Hawkes        & 0.18 & 8.8  & 0.64 (exact)   & ${\sim}2$   & stable, no long-memory \\
    s2p2 (best predictor)  & \textbf{0.32} & 41 & gauge-broken & ---       & over-disperses \\
    NMH / GMH (windowed)   & 0.26 & explode & 1300 / 0.8  & 52--100/s   & explode / Poisson \\
    \textbf{LGM} ($n{=}0.86$) & \textbf{0.29} & \textbf{6.0} & \textbf{0.86 (honest)} & \textbf{2.22/s} & \textbf{calibrated} \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Discussion}
\label{sec:discussion}
% Honest caveats: tails ~2x light; mildly over-reflexive; raw 1s kurtosis/skew
% outlier-dominated (read winsorized / >=5s buckets); no action-conditioning yet.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
